### 2020-08-26 manuscript/paper_BRM.Rmd
Adapt to minor comments by Simon

author: 
  - name          : "Malte Lueken"
    affiliation   : "1"
    corresponding : yes
    address       : "Postal address"
    email         : "malte_lueken@arcor.de"
  - name          : "Simon Kucharsky"
    affiliation   : "1"
  - name          : "Ingmar Visser"
    affiliation   : "1"

@@

Eye-tracking is typically used to study cognitive processes involving attention and information search based on recorded gaze position [@Schulte-Mecklenbeck2017]. Before these processes can be studied, the raw gaze data is classified into events that are distinct in their physiological patterns (e.g., duration), underlying neurological mechanisms, or cognitive functions [@Leigh2015]. Commonly distinguished events are fixations, saccades, smooth pursuits, and post-saccadic oscillations (PSOs). Classifying eye-tracking data reduces their complexity and is the first step towards cognitive interpretation [@Salvucci2000]. The classification is normally done by algorithms, which is considered faster, more objective, and reproducible compared to human coding [@Andersson2017]. @Hein2017 give a comprehensive overview of different classification algorithms [for a structured review on classifying saccades, see also @Stuart2019].  
Gaze data has been frequently classified by probabilistic models. Instead of discrete classification, these models assign probabilities for belonging to an event to each sample. They have the advantage that they learn parameters from the data, can adapt to the task- and individual-specific gaze signals, and can be easily applied online [@Kasneci2014, p. 324].    
One class of probabilistic models that are used in eye movement classification are hidden Markov models (HMMs). They estimate a sequence of hidden states (i.e., a discrete variable that cannot be directly observed) that evolves parallel to the gaze signal. Each gaze sample depends on its corresponding state. Each state depends on the previous but not on earlier states of the sequence [@Zucchini2016].  
In the context of eye movements, HMMs are suitable probabilistic models because the hidden states can be interpreted as eye movement events and gaze data are dependent time series (i.e., one gaze sample depends on the previous). On this basis, several classification algorithms using HMMs have been developed: One instance is described in @Salvucci2000 and combines the HMM with a threshold approach (named "Identification by HMM" [I-HMM]). Samples are first labeled as fixations or saccades, depending on whether their velocity exceeds a threshold, and then reclassified by the HMM. @Pekkanen2017 developed an algorithm that filters the position of gaze samples through naive segmented linear regression (NSLR). The algorithm uses an HMM to parse the resulting segments into fixations, saccades, smooth pursuits, and PSOs based on their velocity and relative angle (named NSLR-HMM). Another version by @Mihali2017 uses a Bayesian HMM to separate microsaccades (short saccades during fixations) from motor noise based on sample velocity (named "Bayesian Microsaccade Detection" [BMD]). Moreover, @Houpt2018 developed a hierarchical approach that describes sample velocity and acceleration through an autoregression (AR) model, computes the regression weights through an HMM, and estimates the number of events with a beta-process (BP) from the data (named BP-AR-HMM).  

@@

In sum, HMMs seem to be a promising method for classifying eye movements. Still, the existing HMM algorithms each have their weaknesses that could be improved upon (and that could explain their inferior performance in comparison to other methods).  
First, I-HMM relies on setting an appropriate threshold, which can distort the results [@Blignaut2009; @Komogortsev2010; @Shic2008]. Second, the current implementation of NSLR-HMM requires human-coded data, which limits its applicability. It also inherits fixed parameters that prevent the algorithm to adapt to the individual- or task-specific signals (contrary to the authors, we argue that this is a disadvantage). Third, BMD limits the classification to microsaccades which are irrelevant in many applications and sometimes even considered as noise [@Duchowski2017]. The opposite problem was observed for BP-AR-HMM: It tends to estimate an unreasonable number of events from the data of which many are considered as noise events. Therefore, the authors suggest using it as an exploratory tool followed by further event classification [@Houpt2018].  
HMMs have another property that has been rarely used in the context of eye movement classification: They can serve as a generative model for gaze data. Only @Mihali2017 used this property by building an explicit generative model for drift and microsaccades.  

@@

# The Algorithm
## Preprocessing
Algorithms require variables that describe gaze data (hereafter called *eye movement metrics*) to classify them into events. Many eye movement metrics have been proposed and used in previous algorithms [for examples, see @Zemblys2018; @Andersson2017]. However, many of them rely on thresholds or window ranges that have to be set by the user [e.g., the distance between the mean position in a 100 ms window before and after each sample, see @Olsson2007]. This can be problematic because such parameters are often set without theoretical justification and they differ substantially between metrics [@Andersson2017]. In gazeHMM, we used velocity, acceleration, and sample-to-sample angle [synonymous to relative or change in angle; @Larsson2013] because they belong to the most basic metrics which do not require parameter settings.  
Theoretically, these three metrics should separate eye movement events clearly. Fixations typically inherit samples with low velocity and acceleration [@Larsson2013]. Due to tremor, we assume that the angle between samples should not follow any direction but a random walk [@Duchowski2017; @Larsson2015]. In contrast, saccade samples usually have a high velocity and acceleration and follow the same direction. PSO samples tend to have moderate velocity and high acceleration since they occur between saccades and low velocity events [@Larsson2013]. They can be specifically distinguished by their change in direction clustered around 180 degrees [@Pekkanen2017]. Importantly, the oscillations depend on the resolution of the gaze recording: Eye-trackers with higher sampling frequency yield more changes in direction and more samples in between those changes. Those samples in between typically follow the same direction. Thus, with high sampling frequencies, PSO samples might also cluster around a sample-to-sample angle of zero with outliers around 180 degrees. Lastly, smooth pursuit samples have a moderate velocity but low acceleration (due to the smoothness) and like saccades they follow the same direction [@Larsson2013; @Leigh2015]. Other algorithms focus exclusively on classifying microsaccades [e.g., @Mihali2017], but as stated earlier, they were not in the scope of gazeHMM.  
The velocity and acceleration signals are computed from the raw gaze position by using a Savitzky-Golay filter [@Savitzky1964; similar to @Nystrom2010]. The sample-to-sample angle is calculated as: $$\alpha(t) = \arctan(\frac{y_{t+1} - y_{t}}{x_{t+1} - x_{t}}) -\arctan(\frac{y_{t} - y_{t-1}}{x_{t} - x_{t-1}}),$$ with $\alpha(t) := \alpha(t) + 2\pi$ for $\alpha(t) < 0$.  
Most of the missing data in eye movement classification are due to blinks. Therefore, the user can indicate which samples should be labeled as blinks (other missing samples are treated as noise). Often, eye-trackers record a few samples with unreasonably high velocity and acceleration before losing the pupil signal when a blink occurs. Since these samples could distort the classification of saccades in the HMM, gazeHMM removes them heuristically. Before classifying the samples, it sets all samples within 50 ms before and after blink samples as missing. The window of 50 ms was rather motivated empirically than theoretically and can bet set to any value the user considers appropriate.

@@

$$Z \sim vonMises(\mu, \kappa)+U(0, 2\pi).$$

@@

## Postprocessing
After classifying gaze samples into states, gazeHMM applies a postprocessing routine to the estimated state sequence. We implemented this routine because constraining the transition probabilities for PSOs to turn into non-saccade events to zero often caused PSOs not to appear in the state sequence at all. Moreover, gazeHMM does not explicitly control the duration of events in the HMM which occasionally led to unreasonably short events. Thus, the postprocessing routine heuristically compensates for such violations. This routine relabels one-sample fixations and smooth pursuits, saccades with a duration below a minimum threshold, and PSOs that follow non-saccade events. Samples are relabeled as the state of the previous event. Finally, samples initially indicated as missing are labeled as noise (including blinks) and event descriptives are computed (e.g., fixation duration).  
The algorithm is implemented in R [version: 3.6.3; @RCoreTeam2020] and uses the packages signal [@Ligges2015] to compute velocity and acceleration signals, depmixS4 [@Visser2010] for the HMM, CircStats [@Lund2018] for the von Mises distribution, and BB [@Varadhan2009] for Barzilai-Borwein spectral projected gradient optimization. The algorithm is available on GitHub (www.github.com/maltelueken/gazeHMM).

@@

## Starting Values
The HMM always started with a uniform distribution to estimate the initial state and state transition probabilities. Random starting values for the estimation of shape, scale, and concentration parameters were generated by gamma distributions with a shape parameter of $\alpha_{start}=3$ and $\beta_{start}=\psi_{true}/2$, with $\psi_{true}$ being the true value of the parameter to be estimated. This setup ensured that the starting values were positive, their distributions were moderately skewed, and the modes of their distributions equaled the true parameter values. Mean parameters of the von Mises distribution always started at their true values.

@@

## Data Analysis
For each parameter, we calculated the root median square proportion deviation [RMdSPD; analogous to root median square percentage errors, see @Hyndman2006] between the true and estimated parameter values: $$\text{RMdSPD} = \sqrt{\text{Med}\left((\frac{\psi_{true}-\psi_{est}}{\psi_{true}})^2\right)}.$$
Even though it was not explicitly mentioned in the preregistration, this measure is only appropriate when $\psi_{true} \ne 0$. This was not the case for some mean parameters of the von Mises distributions. In those cases, we used $\psi_{true} = 2\pi$ instead. We treated $\text{RMdSPD} < 0.1$ as good, $0.1 \le \text{RMdSPD} < 0.5$ as moderate, and $\text{RMdSPD} \ge 0.5$ as bad recovery of a parameter. By taking the median, we reduced the influence of potential outliers in the estimation and using proportions enabled us to compare RMdSPD values across parameters and data sets.  
Additionally, we applied a bivariate linear regression with the estimated parameter values as the dependent and the true parameter values as the independent variable to each parameter that has been varied on an interval in part one. Regression slopes closer to one indicated that the model better captured parameter change. Regression intercepts different from zero reflected a bias in parameter estimation.  
To assess state recovery, we computed Cohen's kappa (for all events taken together, not for each event separately) as a measure of agreement between true and estimated states for each generated data set. Cohen's kappa estimates the agreement between two classifiers accounting for the agreement due to chance. Higher kappa values were interpreted as better model accuracy. We adopted the ranges proposed by @Landis1977 to interpret kappa values.
Models that could not be fitted were excluded from the recovery.

@@

(ref:plot-rmdspd-1-2) RMdSPD between true and estimated parameters of the two-state HMM in part one of the simulation. Labels on the x-axis indicate which true parameters have been manipulated and labels on the y-axis show for which estimated parameter the RMdSPD is displayed. Top facet labels specify in which state the parameters have been varied and right facet labels denote to which state estimated parameters belong.

@@

(ref:plot-regtr-1-2) Regression lines between true and estimated transition probabilities for the two-state HMM in part one. Top facet labels show to and right facet labels show from which state the HMM is moving. Dashed lines refer to perfect recovery.

@@

(ref:plot-rmdspd-1-3) RMdSPD between true and estimated parameters of the three-state HMM in part one of the simulation. Labels on the x-axis indicate which true parameters have been manipulated and labels on the y-axis show for which estimated parameter the RMdSPD is displayed. Top facet labels specify in which state the parameters have been varied and right facet labels denote to which state estimated parameters belong.

@@

(ref:plot-regtr-1-3) Regression lines between true and estimated transition probabilities for the three-state HMM in part one. Top facet labels show to and right facet labels show from which state the HMM is moving. Dashed lines refer to perfect recovery.

@@

The RMdSPDs for the four-state HMM is shown in Figure \@ref(fig:plot-rmdspd-1-4). For estimated transition probabilities and $\alpha_{vel}$ and $\beta_{vel}$ parameters in states one and four, RMdSPDs were between 0.1 and 0.5, suggesting moderate recovery. Also, estimated kappa parameters in state four were often moderately recovered when parameters in states two, three, and four were varied. Otherwise, RMdSPDs were below 0.1, indicating good recovery. Looking at the regression lines between true and estimated parameters, Figures \@ref(fig:plot-regtr-1-4) and \@ref(fig:plot-regresp-1-4) exhibit strong and unbiased relationships. However, there were larger deviations and more outliers than in the previous models, especially for states one and four. Cohen's kappa ranged majorly between 0.6 and 0.9, meaning moderate to almost perfect agreement between true and estimated state sequences (see Figure \@ref(fig:plot-acc-1-4)). Here, some kappa values clustered around 0.25 and zero, which, again, can be interpreted as the result of label switching.    
Overall, simulations in part one demonstrated that the HMM recovered parameters very well when they were manipulated. Deviations from true parameters were mostly small. In the four-state model, estimated transition probabilities for state one and four deviated moderately. Moreover, the HMM estimated state sequences very accurately with decreasing accuracy for the four-state model.

(ref:plot-rmdspd-1-4) RMdSPD between true and estimated parameters of the four-state HMM in part one of the simulation. Labels on the x-axis indicate which true parameters have been manipulated and labels on the y-axis show for which estimated parameter the RMdSPD is displayed. Top facet labels specify in which state the parameters have been varied and right facet labels denote to which state estimated parameters belong.

@@

(ref:plot-regtr-1-4) Regression lines between true and estimated transition probabilities for the four-state HMM in part one. Top facet labels show to and right facet labels show from which state the HMM is moving. Dashed lines refer to perfect recovery.

@@

(ref:plot-rmdspd-2-2) RMdSPD between true and estimated parameters of the two-state HMM in part two of the simulation. Colours indicate different sizes of generated data. Labels on the x-axis indicate for which estimated parameter the RMdSPD is displayed. Top facet labels denote to which state estimated parameters belong.

@@

(ref:plot-rmdspd-2-3) RMdSPD between true and estimated parameters of the three-state HMM in part two of the simulation. Labels on the x-axis indicate for which estimated parameter the RMdSPD is displayed. Top facet labels denote to which state estimated parameters belong.

@@

(ref:plot-rmdspd-2-4) RMdSPD between true and estimated parameters of the three-state HMM in part two of the simulation. Labels on the x-axis indicate for which estimated parameter the RMdSPD is displayed. Top facet labels denote to which state estimated parameters belong.

@@

## Data Analysis
A successful validation of gazeHMM was determined by two approaches: First, we applied gazeHMM with generative models containing 1-5 states to both data sets. The fits of the generative models were compared using Schwarz weights [@Wagenmakers2004], a weighted version of the BIC. They can be interpreted as the probability of a model having generated the data compared to the competing models. For the static condition in the @Andersson2017 data set, we expected the generative model with three states (fixation, saccade, and PSO), and for the dynamic conditions the model with four states (incl. smooth pursuit) to display the highest Schwarz weight. Regarding the @Ehinger2019 data set, we assumed that the one-state model (only fixation) would show the highest weights for both tasks.    
The algorithm was applied separately to every subject, condition, or task. For the @Andersson2017 data set, all generative models were successfully fitted, whereas for the @Ehinger2019 data set, it was only 780 out of 900 models (87%, 60 models per task).  
Second, we compared gazeHMM to other algorithms and human coders. We applied our algorithm with a three-state generative model to the static condition in the @Andersson2017 data set, and with a four-state model to the dynamic conditions. For comparison criteria, we followed @Andersson2017: We calculated the RMSD of event durations and counts between all algorithms and the average of the two human coders. Our results differ slightly from the original study because we excluded two recordings (leading to less events) and calculated the event durations as $Dur(e) = max(\mathbf{t}_{e})-max(\mathbf{t}_{e-1})$, where $\mathbf{t_{e}}$ is the vector of sample time stamps for the event *e*. Cohen's kappa was calculated for each event as the binary agreement between the algorithms and the average of the human coders. Lastly, the overall disagreement indicated which samples were classified differently by the algorithms compared to the average of the human coders across all events. The human coders were compared directly to each other.

>>>>

author: 
  - name          : "Malte Lüken"
    affiliation   : "1"
    corresponding : yes
    address       : "Postal address"
    email         : "malte_lueken@arcor.de"
  - name          : "Šimon Kucharský"
    affiliation   : "1"
  - name          : "Ingmar Visser"
    affiliation   : "1"

@@

Eye-tracking is typically used to study cognitive processes involving attention and information search based on recorded gaze position [@Schulte-Mecklenbeck2017]. Before these processes can be studied, the raw gaze data is classified into events that are distinct in their physiological patterns (e.g., duration), underlying neurological mechanisms, or cognitive functions [@Leigh2015]. Basic events are fixations, saccades, smooth pursuits, and post-saccadic oscillations (PSOs). Classifying raw eye-tracking data into these events reduces their complexity and is usually the first step towards cognitive interpretation [@Salvucci2000]. The classification is typically done by algorithms, which is considered faster, more objective, and reproducible compared to human coding [@Andersson2017]. @Hein2017 give a comprehensive overview of different classification algorithms [for a structured review on classifying saccades, see also @Stuart2019].  
Gaze data has been frequently classified by probabilistic models. Compared to discrete (hard) classification, these models compute the probability of each event for every sample of the raw eye-tracking data. They have the advantage that they learn parameters from the data, can adapt to the task- and individual-specific gaze signals, and can be easily applied online [@Kasneci2014, p. 324].    
One class of probabilistic models that are used in eye movement classification are hidden Markov models (HMMs). HMMs estimate a sequence of hidden states (i.e., a discrete variable that cannot be directly observed) that evolves parallel to the gaze signal. Each gaze sample depends on its corresponding state. Each state depends on the previous but not on earlier states of the sequence [@Zucchini2016].  
In the context of eye movements, HMMs are suitable probabilistic models because the hidden states can be interpreted as eye movement events and gaze data are dependent time series (i.e., one gaze sample depends on the previous). Further, HMMs can be viewed as unsupervised models that can learn the hidden states and parameters of the emission process from the observed data alone, and as such do not in principle need training labeled data as it would be the case for supervised machine learning. On this basis, several classification algorithms using HMMs have been developed: One instance is described in @Salvucci2000 and combines the HMM with a fixed threshold approach (named "Identification by HMM" [I-HMM]). Samples are first labeled as fixations or saccades, depending on whether their velocity exceeds a threshold, and then reclassified by the HMM. @Pekkanen2017 developed an algorithm that filters the position of gaze samples through naive segmented linear regression (NSLR). The algorithm uses an HMM to parse the resulting segments into fixations, saccades, smooth pursuits, and PSOs based on their velocity and relative angle (named NSLR-HMM). Another version by @Mihali2017 uses a Bayesian HMM to separate microsaccades (short saccades during fixations) from motor noise based on sample velocity (named "Bayesian Microsaccade Detection" [BMD]). Moreover, @Houpt2018 developed a hierarchical approach that describes sample velocity and acceleration through an autoregression (AR) model, computes the regression weights through an HMM, and estimates the number of events with a beta-process (BP) from the data (named BP-AR-HMM).  

@@

In sum, HMMs seem to be a promising method for classifying eye movements. However, the existing HMM algorithms each have their weaknesses that could be improved upon.  
First, I-HMM relies on setting an appropriate threshold to determine the initial classification, which can distort the results [@Blignaut2009; @Komogortsev2010; @Shic2008]. Second, the current implementation of NSLR-HMM requires human-coded data, which limits its applicability. It also inherits fixed parameters that prevent the algorithm to adapt to the individual- or task-specific signals (contrary to the authors, we argue that this is a disadvantage). Third, BMD limits the classification to microsaccades which are irrelevant in many applications and sometimes even considered as noise [@Duchowski2017]. The opposite problem was observed for BP-AR-HMM: It tends to estimate an unreasonable number of events from the data of which many are considered as noise events. Therefore, the authors suggest using it as an exploratory tool followed by further event classification [@Houpt2018].  
HMMs have another property that has been rarely used in the context of eye movement classification: They can serve as a generative model for gaze data. To our knowledge, only @Mihali2017 used this property by building an explicit generative model for drift and microsaccades.  

@@

# The Algorithm
## Preprocessing
Algorithms require variables that describe gaze data (hereafter called *eye movement metrics*) to classify them into events. Many eye movement metrics have been proposed and used in previous algorithms [for examples, see @Zemblys2018; @Andersson2017]. However, many of them rely on thresholds or window ranges that have to be set by the user [e.g., the distance between the mean position in a 100 ms window before and after each sample, see @Olsson2007]. This can be problematic because such parameters are often set without theoretical justification and differ substantially between metrics or heavily depend on the eye-tracker's sampling frequency [@Andersson2017]. In gazeHMM, we used velocity, acceleration, and sample-to-sample angle [synonymous to relative or change in angle; @Larsson2013] because they belong to the most basic metrics which do not require parameter settings.  
Theoretically, these three metrics should separate eye movement events clearly. Fixations typically inherit samples with low velocity and acceleration [@Larsson2013]. Due to tremor, we assume that the angle between samples should not follow any direction but a random walk [@Duchowski2017; @Larsson2015]. In contrast, saccade samples usually have a high velocity and acceleration and roughly follow the same direction. PSO samples tend to have moderate velocity and high acceleration since they occur between saccades and low velocity events [@Larsson2013]. They can be specifically distinguished by their change in direction clustered around 180 degrees [@Pekkanen2017]. Importantly, the oscillations depend on the resolution of the gaze recording: Eye-trackers with higher sampling frequency yield more changes in direction and more samples in between those changes. Those samples in between typically follow the same direction. Thus, with high sampling frequencies, PSO samples might also cluster around a sample-to-sample angle of zero with outliers around 180 degrees. Lastly, smooth pursuit samples have a moderate velocity but low acceleration (due to the smoothness) and like saccades they follow the same direction [@Larsson2013; @Leigh2015]. Other algorithms focus exclusively on classifying microsaccades [e.g., @Mihali2017], but as stated earlier, these events were not in the scope of gazeHMM.  
The velocity and acceleration signals are computed from the raw gaze position by using a Savitzky-Golay filter [@Savitzky1964; similar to @Nystrom2010]. The sample-to-sample angle is calculated as: $$\alpha(t) = \arctan\left(\frac{y_{t+1} - y_{t}}{x_{t+1} - x_{t}}\right) -\arctan\left(\frac{y_{t} - y_{t-1}}{x_{t} - x_{t-1}}\right),$$ with $\alpha(t) := \alpha(t) + 2\pi$ for $\alpha(t) < 0$, and is therefore bound between 0 and $2\pi$.  
Most of the missing data in eye movement classification are due to blinks. Therefore, the user can indicate which samples should be labeled as blinks (other missing samples are treated as noise). Often, eye-trackers record a few samples with unreasonably high velocity and acceleration before losing the pupil signal when a blink occurs. Since these samples could distort the classification of saccades in the HMM, gazeHMM removes them heuristically. Before classifying the samples, it sets all samples within 50 ms before and after blink samples as missing. The window of 50 ms was rather motivated empirically than theoretically and can bet set to any value the user considers appropriate.

@@

$$\begin{aligned}
Z \mid fixation &\sim U(0, 2\pi) \\
Z \mid saccade &\sim von\ Mises(\mu, \kappa) \\
Z \mid PSO &\sim von\ Mises(\mu, \kappa) \\
Z \mid smooth\ pursuit &\sim von\ Mises(\mu, \kappa)
\end{aligned}.$$

@@

## Postprocessing
After classifying gaze samples into states, gazeHMM applies a postprocessing routine to the estimated state sequence. We implemented this routine because constraining the transition probabilities for PSOs to turn into non-saccade events to zero often caused PSOs not to appear in the state sequence at all. Moreover, gazeHMM does not explicitly control the duration of events in the HMM which occasionally led to unreasonably short events. Thus, the postprocessing routine heuristically compensates for such violations. This routine relabels one-sample fixations and smooth pursuits, saccades with a duration below a minimum threshold, and PSOs that follow non-saccade events. Samples are relabeled as the state of the previous event. Finally, samples initially indicated as missing are labeled as noise (including blinks) and event descriptives are computed (e.g., fixation duration).  
The algorithm is implemented in R [version: 3.6.3; @RCoreTeam2020] and uses the packages \texttt{signal} [@Ligges2015] to compute velocity and acceleration signals, \texttt{depmixS4} [@Visser2010] for the HMM, \texttt{CircStats} [@Lund2018] for the von Mises distribution, and \texttt{BB} [@Varadhan2009] for Barzilai-Borwein spectral projected gradient optimization. The algorithm is available on GitHub (www.github.com/maltelueken/gazeHMM).

@@

## Starting Values
The HMM always started with a uniform distribution for the initial state and state transition probabilities. Random starting values for the estimation of shape, scale, and concentration parameters were generated by gamma distributions with a shape parameter of $\alpha_{start}=3$ and $\beta_{start;i}=\theta_i/2$, with $\theta_i$ being the true value of the parameter to be estimated in simulation $i\in(1,\dots,I)$. This setup ensured that the starting values were positive, their distributions were moderately skewed, and the modes of their distributions equaled the true parameter values. Mean parameters of the von Mises distribution always started at their true values.

@@

## Data Analysis
For each parameter, we calculated the root median square proportion deviation [RMdSPD; analogous to root median square percentage errors, see @Hyndman2006] between the true and estimated parameter values:
$$
\begin{aligned}
\text{RMdSPD} & = \sqrt{\text{Median}(\epsilon_1^2,\dots,\epsilon_I^2)} \\ 
\epsilon_i^2 & = \left(\frac{\hat{\theta}_i - \theta_i}{\theta_i}\right)^2, 
\end{aligned}
$$
where $\theta_i$ is the true parameter value and $\hat{\theta}_i$ is the estimated parameter value for simulation $i\in(1,\dots,I)$, respectively. Even though it was not explicitly mentioned in the preregistration, this measure is only appropriate when $\theta_i \ne 0$. This was not the case for some mean parameters of the von Mises distributions. In those cases, we used $\theta_i = 2\pi$ instead. We treated $\text{RMdSPD} < 0.1$ as good, $0.1 \le \text{RMdSPD} < 0.5$ as moderate, and $\text{RMdSPD} \ge 0.5$ as bad recovery of a parameter. By taking the median, we reduced the influence of potential outliers in the estimation and using proportions enabled us to compare RMdSPD values across parameters and data sets.  
Additionally, we applied a bivariate linear regression with the estimated parameter values as the dependent and the true parameter values as the independent variable to each parameter that has been varied on an interval in part one. Regression slopes closer to one indicated that the model better captured parameter change. Regression intercepts different from zero reflected a bias in parameter estimation.  
To assess state recovery, we computed Cohen's kappa (for all events taken together, not for each event separately) as a measure of agreement between true and estimated states for each generated data set. Cohen's kappa estimates the agreement between two classifiers accounting for the agreement due to chance. Higher kappa values were interpreted as better model accuracy. We adopted the ranges proposed by @Landis1977 to interpret kappa values.
Models that could not be fitted were excluded from the recovery.

@@

(ref:plot-rmdspd-1-2) RMdSPD between true and estimated parameters of the two-state HMM in part one of the simulation. Labels on the x-axis indicate which true parameters have been manipulated and labels on the y-axis show for which estimated parameter the RMdSPD is displayed. Top facet labels specify in which state the parameters have been varied and right facet labels denote to which state estimated parameters belong. $\rho_i$ is the initial probability for state *i* (indicated by the right facet label), $a_{ij}$ is the probability to transition from state *i* to state *j*, $\alpha$ and $\beta$ are the shape and scale parameters of the gamma distributions, and $\mu$ and $\kappa$ are the mean and concentration parameter of the von Mises distribution.

@@

(ref:plot-regtr-1-2) Regression lines between true and estimated transition probabilities for the two-state HMM in part one. Right facet labels show *from* and top facet labels show *to* which state the HMM is moving. Dashed lines refer to perfect recovery.

@@

(ref:plot-rmdspd-1-3) RMdSPD between true and estimated parameters of the three-state HMM in part one of the simulation. Labels on the x-axis indicate which true parameters have been manipulated and labels on the y-axis show for which estimated parameter the RMdSPD is displayed. Top facet labels specify in which state the parameters have been varied and right facet labels denote to which state estimated parameters belong. $\rho_i$ is the initial probability for state *i* (indicated by the right facet label), $a_{ij}$ is the probability to transition from state *i* to state *j*, $\alpha$ and $\beta$ are the shape and scale parameters of the gamma distributions, and $\mu$ and $\kappa$ are the mean and concentration parameter of the von Mises distribution.

@@

(ref:plot-regtr-1-3) Regression lines between true and estimated transition probabilities for the three-state HMM in part one. Right facet labels show *from* and top facet labels show *to* which state the HMM is moving. Dashed lines refer to perfect recovery.

@@

The RMdSPDs for the four-state HMM is shown in Figure \@ref(fig:plot-rmdspd-1-4). For estimated transition probabilities and $\alpha_{vel}$ and $\beta_{vel}$ parameters in states one and four, RMdSPDs were between 0.1 and 0.5, suggesting moderate recovery. Also, estimated kappa parameters in state four were often moderately recovered when parameters in states two, three, and four were varied. Otherwise, RMdSPDs were below 0.1, indicating good recovery. Looking at Figures \@ref(fig:plot-regtr-1-4) and \@ref(fig:plot-regresp-1-4), the regression lines between true and estimated parameters exhibit strong and unbiased relationships. However, there were larger deviations and more outliers than in the previous models, especially for states one and four. Cohen's kappa ranged mostly between 0.6 and 0.9, meaning moderate to almost perfect agreement between true and estimated state sequences (see Figure \@ref(fig:plot-acc-1-4)). Here, some kappa values clustered around 0.25 and zero, which, again, suggests label switching.    
Overall, simulations in part one demonstrated that the HMM recovered parameters very well when they were manipulated. Deviations from true parameters were mostly small. In the four-state model, estimated transition probabilities for state one and four deviated moderately. Moreover, the HMM estimated state sequences very accurately with decreasing accuracy for the four-state model.

(ref:plot-rmdspd-1-4) RMdSPD between true and estimated parameters of the four-state HMM in part one of the simulation. Labels on the x-axis indicate which true parameters have been manipulated and labels on the y-axis show for which estimated parameter the RMdSPD is displayed. Top facet labels specify in which state the parameters have been varied and right facet labels denote to which state estimated parameters belong. $\rho_i$ is the initial probability for state *i* (indicated by the right facet label), $a_{ij}$ is the probability to transition from state *i* to state *j*, $\alpha$ and $\beta$ are the shape and scale parameters of the gamma distributions, and $\mu$ and $\kappa$ are the mean and concentration parameter of the von Mises distribution.

@@

(ref:plot-regtr-1-4) Regression lines between true and estimated transition probabilities for the four-state HMM in part one. Right facet labels show *from* and top facet labels show *to* which state the HMM is moving. Dashed lines refer to perfect recovery.

@@

(ref:plot-rmdspd-2-2) RMdSPD between true and estimated parameters of the two-state HMM in part two of the simulation. Colours indicate different sizes of generated data. Labels on the x-axis indicate for which estimated parameter the RMdSPD is displayed. Top facet labels denote to which state estimated parameters belong. $\rho_i$ is the initial probability for state *i* (indicated by the top facet label), $a_{ij}$ is the probability to transition from state *i* to state *j*, $\alpha$ and $\beta$ are the shape and scale parameters of the gamma distributions, and $\mu$ and $\kappa$ are the mean and concentration parameter of the von Mises distribution.

@@

(ref:plot-rmdspd-2-3) RMdSPD between true and estimated parameters of the three-state HMM in part two of the simulation. Labels on the x-axis indicate for which estimated parameter the RMdSPD is displayed. Top facet labels denote to which state estimated parameters belong. $\rho_i$ is the initial probability for state *i* (indicated by the top facet label), $a_{ij}$ is the probability to transition from state *i* to state *j*, $\alpha$ and $\beta$ are the shape and scale parameters of the gamma distributions, and $\mu$ and $\kappa$ are the mean and concentration parameter of the von Mises distribution.

@@

(ref:plot-rmdspd-2-4) RMdSPD between true and estimated parameters of the three-state HMM in part two of the simulation. Labels on the x-axis indicate for which estimated parameter the RMdSPD is displayed. Top facet labels denote to which state estimated parameters belong. $\rho_i$ is the initial probability for state *i* (indicated by the top facet label), $a_{ij}$ is the probability to transition from state *i* to state *j*, $\alpha$ and $\beta$ are the shape and scale parameters of the gamma distributions, and $\mu$ and $\kappa$ are the mean and concentration parameter of the von Mises distribution.

@@

## Data Analysis
A successful validation of gazeHMM was determined by two approaches: First, we applied gazeHMM with generative models containing 1-5 states to both data sets. The fits of the generative models were compared using Schwarz weights [@Wagenmakers2004], a conversion of the BIC [@Schwarz1978] into model weights. They can be interpreted as the probability of a model having generated the data compared to the competing models. For the static condition in the @Andersson2017 data set, we expected the generative model with three states (fixation, saccade, and PSO), and for the dynamic conditions the model with four states (incl. smooth pursuit) to display the highest Schwarz weight. Regarding the @Ehinger2019 data set, we assumed that the one-state model (only fixation) would show the highest weights for both tasks.    
The algorithm was applied separately to every subject, condition, or task. For the @Andersson2017 data set, all generative models were successfully fitted, whereas for the @Ehinger2019 data set, it was only 780 out of 900 models (87%, 60 models per task).  
Second, we compared gazeHMM to other algorithms and human coders. We applied our algorithm with a three-state generative model to the static condition in the @Andersson2017 data set, and with a four-state model to the dynamic conditions. For comparison criteria, we followed @Andersson2017: We calculated the RMSD of event durations and counts between all algorithms and the average of the two human coders. Our results differ slightly from the original study because we excluded two recordings (leading to less events) and calculated the event durations as $Dur(e) = max(\mathbf{t}_{e})-max(\mathbf{t}_{e-1})$, where $\mathbf{t_{e}}$ is the vector of sample time stamps for the event *e*. Cohen's kappa was calculated for each event as the binary agreement between the algorithms and the average of the human coders. Lastly, the overall disagreement indicated which samples were classified differently by the algorithms compared to the average of the human coders across all events. The human coders were compared directly to each other.
